---
title: "Fig 1"
author: "Ning Chin"
date: "3/24/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

Setup
```{r setup, warning = FALSE, message = FALSE}
library(phyloseq)
library(tidyr)
library(vegan)
library(ggplot2)
library(cowplot)
library(ggdendro)
library(ggrepel)

set.seed(7)

set.color <- as.character(c("#DD1717", "#000249", "#FF4949", "#0F4392", 
                            "#FCCDE5", "#8DD3C7", "#FFFFB3", "#80B1D3", "#FB8072", "#BC80BD",
                            "#FFED6F", "#D9D9D9", "#BEBADA", "#CCEBC5", "#FDB462", "#B3DE69"))
names(set.color) <- c("NEG", "POS", "prev", "posv",
                      "Actinobacteria", "Bacteroidetes", "Cyanobacteria", "Elusimicrobia", "Fibrobacteres", "Firmicutes",
                      "Proteobacteria", "Spirochaetes", "Tenericutes", "TM7", "Verrucomicrobia", "WPS-2")

theme_set(theme_grey()+
            theme(panel.grid = element_blank(),
                  panel.background = element_blank(),
                  axis.title = element_text(size = 10),
                  axis.text = element_text(size = 6),
                  legend.title = element_text(size = 8, face = "bold"),
                  legend.text = element_text(size = 6),
                  legend.key = element_blank(),
                  legend.key.size = unit(10, "pt")))
```

## General filtering
```{r filt}
filterfunc <- function(psobj, prevpercent) {
  ps0 <- subset_taxa(psobj, !is.na(Phylum))
  ps1 <- tax_glom(ps0, "Genus", NArm = FALSE)
  prevdf <- apply(otu_table(ps1), 2, FUN = function(x){sum(x > 0)})
  prevalenceThreshold <-  prevpercent * nsamples(ps1)
  keepTaxa <- names(prevdf)[(prevdf >= prevalenceThreshold)]
  ps2 <- prune_taxa(keepTaxa, ps1)
  return(ps2)
}

ps0 <- readRDS("../phyloseq_GG_tree_190304.RData")
ps <- filterfunc(ps0, 0.05)
ps.ra <- transform_sample_counts(ps, function(x) x / sum(x) * 100)
ps.clr <- transform_sample_counts(ps, function(x) log(x + 1) - mean(log(x + 1)))
mtdata <- data.frame(sample_data(ps))
mtdata$AnimalID <- as.character(mtdata$AnimalID)
```

## Fig 1A. Barplot at phylum level
```{r 1a}
phy <- tax_glom(ps.ra, "Phylum")
phydf <- psmelt(phy)

# get order from clustering using Euclidean/Aitchinson distance
bac <- data.frame(otu_table(ps.clr))
row.clus <- hclust(dist(bac), method = "complete")
dend <- as.dendrogram(row.clus)
ddata <- dendro_data(dend, type = "rectangle")
ddata <- data.frame(segment(ddata))

ddata.label <- dplyr::left_join(data.frame(AnimalID = row.clus$labels[row.clus$order]), mtdata[,c("AnimalID", "CMV")], by = "AnimalID")
ddata.label$x <- 1:nrow(ddata.label)

p1 <- ggplot(ddata, aes(x = x, y = y)) +
  geom_segment(aes(xend = xend, yend = yend), size = 0.2) +
  scale_x_continuous(expand = c(0.01, 0.01)) +
  geom_point(data = ddata.label, aes(y = -3.5, color = CMV), size = 1.2, shape = 15, alpha = 0.8) +
  scale_color_manual(values = set.color, limits = force, name = "RhCMV") +
  theme(axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = c(0.8, 1.1),
        legend.direction = "horizontal",
        legend.text = element_text(size = 7),
        legend.margin = margin(b = -10),
        plot.margin = margin(t = 5, b = 2, r = 5, l = 5))
p1

phydf$Sample <- factor(phydf$Sample, levels = row.clus$labels[row.clus$order])
phydf$Phylum <- gsub("p__*", "", phydf$Phylum)

p2 <- ggplot(phydf, aes(x = Sample, y = Abundance)) +
  geom_bar(stat = "identity", position = "stack", color = "lightgrey", aes(fill = Phylum), size = 1e-6) +
  ylab("Relative Abundance (%)") +
  scale_fill_manual(values = set.color, limits = force) +
  guides(fill = guide_legend(nrow = 4)) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_text(margin = margin(t = -7)),
        axis.title.y = element_text(margin = margin(r = -2)),
        legend.position = "bottom",
        legend.key.size = unit(10, "pt"),
        legend.margin = margin(t = -10),
        plot.margin = margin(t = -10, r = 5, l = 5))

fig1a <- plot_grid(p1, p2,
                   ncol = 1, align = "v", axis = "lr", rel_heights = c(1, 3.5))
fig1a
```

## Alpha diversities
```{r alpha.div}
Diversity <- diversity(otu_table(ps), index = "shannon")
Richness <- specnumber(otu_table(ps))
Evenness <- Diversity/log(Richness)

df <- data.frame(sample_data(ps), Diversity, Richness, Evenness)
df.long <- gather(df, method, value, Diversity:Evenness)

ggplot(df.long, aes(x = CMV, y = value, fill = CMV)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  facet_wrap(~method, scales = "free")
```

Shannon diversity p-value = `r round(wilcox.test(df$Diversity[df$CMV == "NEG"], df$Diversity[df$CMV == "POS"])$p.value, 3)`.\
Evenness p-value = `r round(wilcox.test(df$Evenness[df$CMV == "NEG"], df$Evenness[df$CMV == "POS"])$p.value, 3)`.\
Richness p-value = `r round(wilcox.test(df$Richness[df$CMV == "NEG"], df$Richness[df$CMV == "POS"])$p.value, 3)`.

## Fig 1B. PCA / Aitchison distance
```{r 1b}
# Aitchison = PCA of CLR transformed counts
bac <- data.frame(otu_table(ps.clr))
colnames(bac) <- data.frame(tax_table(ps.clr))$Genus

pca <- prcomp(bac, scale. = TRUE)
out <- cbind(mtdata, data.frame(pca$x))
percentage <- round(pca$sdev^2 / sum(pca$sdev^2) * 100, 2)
eu <- dist(bac)

# beta-dispersion p-value:
permutest(betadisper(eu, mtdata$CMV))$tab$`Pr(>F)`[1]

p.label <- paste0("P = ", round(adonis(eu ~ CMV, data = out, method = "eu")$aov.tab$`Pr(>F)`[1],3))

fig1b <- ggplot(out, aes(x = PC1, y = PC2)) +
  geom_point(size = 2, alpha = 0.8, aes(color = CMV)) +
  labs(x = paste("PC1 (", percentage[1], "%)", sep = ""),
       y = paste("PC2 (", percentage[2], "%)", sep = "")) +
  scale_color_manual(values = set.color, name = "RhCMV", limits = force) +
  theme(panel.border = element_rect(fill = NA, colour = "black", size = 0.2),
        axis.title.y = element_text(margin = margin(r = -2)),
        legend.position = "top",
        legend.margin = margin(b = -10),
        legend.text = element_text(size = 7),
        aspect.ratio = 1) +
  annotate("text",  x = -Inf, y = Inf, fontface = "italic",
           vjust = 1.4, hjust = -0.2, 
           size = 2,
           label = p.label)
fig1b
```

Sex did not effect clustering:
PERMANOVA p-value based on Sex = `r round(adonis(eu ~ Sex, data = mtdata)$aov.tab$'Pr(>F)'[1], 3)`.

## Put all plots together
```{r fig1, fig.show = "hide"}
plot_grid(fig1a, fig1b, nrow = 1, rel_widths = c(1.2, 1), labels = c("A", "B"), label_size = 16, hjust = -0.1, vjust = 1.1)

ggsave("211028_fig1_CLR.pdf", width = 6.6, height = 3.2, units = "in", device = "pdf")

sessionInfo()
```