---
title: "Fig 2"
author: "Ning Chin"
date: "10/1/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

Setup
```{r setup, warning = FALSE, message = FALSE}
library(phyloseq)
library(ggplot2)
library(edgeR)
library(tidyr)
library(cowplot)
library(glmnet)
library(pROC)
library(kableExtra)
library(vegan)

set.seed(7)

set.color <- as.character(c("#DD1717", "#000249", "#FF4949", "#0F4392", 
                            "#FCCDE5", "#8DD3C7", "#FFFFB3", "#80B1D3", "#FB8072", "#BC80BD",
                            "#FFED6F", "#D9D9D9", "#BEBADA", "#CCEBC5", "#FDB462", "#B3DE69"))
names(set.color) <- c("NEG", "POS", "prev", "posv",
                      "Actinobacteria", "Bacteroidetes", "Cyanobacteria", "Elusimicrobia", "Fibrobacteres", "Firmicutes",
                      "Proteobacteria", "Spirochaetes", "Tenericutes", "TM7", "Verrucomicrobia", "WPS-2")

theme_set(theme_grey()+
            theme(panel.grid = element_blank(),
                  panel.background = element_blank(),
                  panel.border = element_rect(fill = NA, colour = "black", size = 0.2),
                  axis.title = element_text(size = 10),
                  axis.text = element_text(size = 6),
                  axis.line = element_line(color = "black", size = 0.2),
                  legend.title = element_text(size = 8, face = "bold"),
                  legend.text = element_text(size = 6),
                  legend.key = element_blank(),
                  legend.key.size = unit(10, "pt"),
                  plot.margin = unit(c(5, 5, 5, 5), "pt")))
```

## Filtering
```{r filt}
filterfunc <- function(psobj, prevpercent, maxpercent) {
  ps0 <- subset_taxa(psobj, !is.na(Phylum))
  ps1 <- tax_glom(ps0, "Genus", NArm = TRUE)
  prevalence <- apply(otu_table(ps1), 2, function(x) sum(x > 0)/nsamples(ps1))
  maxpercentage <- apply(otu_table(ps1), 1, function(x) x/sum(x)*100)
  maxpercentage <- apply(maxpercentage, 1, function(x) max(x))
  keepTaxa <- names(prevalence)[(prevalence >= prevpercent & maxpercentage > maxpercent)]
  ps2 <- prune_taxa(keepTaxa, ps1)
  return(ps2)
}

ps0 <- readRDS("../phyloseq_GG_tree_190304.RData")
dim(tax_table(ps0))
ps <- filterfunc(ps0, 0.05, 0.01)
ps.ra <- transform_sample_counts(ps, function(x) x / sum(x) * 100)
dim(tax_table(ps))
mtdata <- read.csv("../metadata.csv", stringsAsFactors = TRUE)
taxa <- data.frame(tax_table(ps))

# CMV38 data
vac0 <- readRDS("../../CMV38/phyloseq_GG_tree.RData")
vac <- filterfunc(vac0, 0.05, 0.01)
vac.ra <- transform_sample_counts(vac, function(x) x / sum(x) * 100)
vacmtdata <- read.csv("../../CMV38/metadata.csv")
vacmtdata$Timepoint <- factor(vacmtdata$Timepoint, levels = c("prev", "posv"))
```

```{r limma-voom}
limmafunc <- function(psobj) {
  m <- t(otu_table(psobj))
  d <- DGEList(counts = m, genes = data.frame(as(tax_table(psobj),"matrix"), stringsAsFactors = FALSE))
  mm <- model.matrix(~ CMV, data = data.frame(sample_data(psobj), stringsAsFactors = FALSE))
  y <- voom(d, mm, plot = TRUE)
  fit <- lmFit(y, mm)
  contr <- makeContrasts(CMVPOS, levels = colnames(coef(fit)))
  tmp <- contrasts.fit(fit, contr)
  tmp <- eBayes(tmp)
  tmp2 <- topTable(tmp, coef = 1, sort.by = "P", n = Inf)
  return(tmp2)
}

lim <- limmafunc(ps)
lim$Phylum <- gsub("p__", "", lim$Phylum)
lim$Genus <- gsub("g__", "", lim$Genus)
```


## Fig 2A. Waterfall plot for significantly changed taxa
```{r 2a}
sigp <- lim[lim$P.Value < 0.05,]
sigp$Genus <- factor(sigp$Genus, levels = sigp$Genus[order(sigp$logFC, decreasing = TRUE)])
sigp$sig <- ifelse(sigp$adj.P.Val < 0.1, "*", "")
sigp$sig.y.val <- ifelse(sigp$logFC > 0, sigp$logFC, 0)

kable(sigp, row.names = 0) %>% kable_styling()

fig2a <- ggplot(sigp, aes(x = Genus, y = logFC, fill = Phylum)) +
  geom_col(position = "dodge", color = "black", size = 0.2) +
  geom_text(aes(y = sig.y.val), label = sigp$sig, size = 7) +
  labs(y = "log2 FC") +
  ylim(-2.3, 2.7) +
  scale_fill_manual(values = set.color, limits = force) +
  theme(panel.border = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 8, face = "italic"),
        legend.position = "none") +
  annotate("text",  x = Inf, y = Inf, fontface = "italic",
           vjust = 1.2, hjust = 1.2, 
           size = 2,
           label = "* adj. P < 0.1")
fig2a

# make legend that include all 6 phylum
fig2.legend <- ggplot(lim, aes(x = Genus, y = logFC, fill = Phylum)) +
  geom_col(color = "black", size = 0.2) +
  scale_fill_manual(values = set.color, limits = force) +
  guides(fill = guide_legend(nrow = 2)) +
  theme(legend.position = "bottom",
        legend.margin = margin(-10, 0, -5, 0))
```


## Run elastic net logistic regression for feature selection
```{r}
# use CUR01 as training set and CMV38 as testing set
# keep common genus only
keep <- intersect(data.frame(tax_table(ps), stringsAsFactors = FALSE)$Genus, data.frame(tax_table(vac), stringsAsFactors = FALSE)$Genus)

# train set from CUR01
bac.train <- data.frame(otu_table(ps.ra))
colnames(bac.train) <- data.frame(tax_table(ps), stringsAsFactors = FALSE)$Genus
x.train <- scale(bac.train[,keep])
y.train <- mtdata$CMV

# test set from CMV38
bac.test <- data.frame(otu_table(vac.ra))
colnames(bac.test) <- data.frame(tax_table(vac), stringsAsFactors = FALSE)$Genus
x.test <- scale(bac.test[,keep])
y.test <- as.factor(ifelse(vacmtdata$Timepoint == "prev", "NEG", "POS"))

# test a range of alpha and select one with minimum cross-validation error
# alphas to be tested (0 = ridge, 1 = lasso)
results <- data.frame(alpha = seq(0, 1, 0.05), cvm = 0)
list.of.fits <- list()

for (i in 1:nrow(results)){
  # perform 10-fold cross-validation
  fit <- cv.glmnet(x.train, y.train, family = "binomial", type.measure = "class", alpha = results$alpha[i], standardize = FALSE, nfolds = 10)
  # save results to data.frame
  results$cvm[i] <- fit$cvm[fit$lambda == fit$lambda.1se]
  # save model in case we need it later
  list.of.fits[[paste0("alpha", results$alpha[i])]] <- fit
}

# extract the optimal cross-validated model
op.fit <- list.of.fits[[paste0("alpha", results$alpha[results$cvm == min(results$cvm)])]]

# plot cross-validation results
plot(op.fit)
```


## Fig 2B. Plot coefficient of ENR selected features
```{r}
enr.res <- coef(op.fit)
enr.res <- data.frame(Genus = enr.res@Dimnames[[1]][enr.res@i + 1], Coefficient = enr.res@x)[-1,]
enr.res <- dplyr::right_join(taxa, enr.res, by = "Genus")
enr.res$Genus <- gsub("g__", "", enr.res$Genus)
enr.res$Genus <- factor(enr.res$Genus, levels = enr.res$Genus[order(enr.res$Coefficient, decreasing = TRUE)])
enr.res$Phylum <- gsub("p__", "", enr.res$Phylum)
kable(enr.res) %>% kable_styling()

fig2b <- ggplot(enr.res, aes(x = Genus, y = Coefficient, fill = Phylum)) +
  geom_col(position = "dodge", color = "black", size = 0.2) +
  scale_fill_manual(values = set.color, limits = force) +
  scale_y_continuous(limit = c(-1.2, 0.4), expand = expansion(mult = c(0,0), add = c(0.01, 0.02))) +
  theme(panel.border = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 8, face = "italic"),
        legend.position = "none")
fig2b

# get cmv_score - save for use later
cmv_score <- predict(op.fit, s = op.fit$lambda.1se, newx = x.train, type = "response")
#saveRDS(cmv_score, "cmv_score.RData")
```


## Fig 2C. PCA for significantly changed taxa in CUR01
```{r}
# subset what we're interested in
sigbac <- paste0("g__", enr.res$Genus)
ps.sig <- subset_taxa(ps, Genus %in% sigbac)

# get log transformed counts
ps.log <- transform_sample_counts(ps.sig, function(x) log2(x + 1))

logbac <- data.frame(otu_table(ps.log))
colnames(logbac) <- data.frame(as(tax_table(ps.log), "matrix"), stringsAsFactors = FALSE)$Genus

pca <- prcomp(logbac, scale. = TRUE)
out <- cbind(mtdata, as.data.frame(pca$x))
percentage <- round(pca$sdev^2 / sum(pca$sdev^2) * 100, 2)

eu <- dist(logbac)

fig2c <- ggplot(out, aes(x = PC1, y = PC2)) +
  geom_point(size = 2, alpha = 0.8, aes(color = CMV)) +
  labs(x = paste("PC1 (", percentage[1], "%)", sep = ""),
       y = paste("PC2 (", percentage[2], "%)", sep = ""),
       title = "Training set",
       subtitle = "RhCMV- vs. RhCMV+") +
  scale_color_manual(values = set.color, limits = force, name = "RhCMV") +
  theme(legend.position = c(1, 1),
        legend.justification = c(1, 1),
        legend.background = element_rect(color = "black", size = 0.2),
        legend.spacing.y = unit(1, "pt"),
        legend.margin = margin(0, 2, 0, 2),
        aspect.ratio = 1,
        plot.title = element_text(size = 12, face = "bold", margin = margin(0, 0, 0, 0)),
        plot.subtitle = element_text(size = 8, face = "italic", margin = margin(0, 0, -6, 0), hjust = 1, vjust = 6)) +
  annotate("text",  x = -Inf, y = Inf, fontface = "italic",
           vjust = 1.4, hjust = -0.2, size = 2,
           label = paste0("P = ",
                    round(adonis(eu ~ CMV, data = out, method = "eu")$aov.tab$`Pr(>F)`[1],3)))
fig2c

# beta-disper p-value
round(permutest(betadisper(eu, mtdata$CMV))$tab$'Pr(>F)'[1], 3)
```


## Fig 2D. AUC curve for ENR
```{r}
# apply on test set, graph AUC curve with confidence interval of sensitivities
predicted <- predict(op.fit, s = op.fit$lambda.1se, newx = x.test, type = "response")
enr.auc <- roc(y.test, as.numeric(predicted), ci = TRUE)
enr.obj <- ci.se(enr.auc)

# extract auc info for ggplot
enr.df <- coords(enr.auc, "all", transpose = FALSE)
enr.df <- enr.df[rev(seq(nrow(enr.df))),]

# extract confidence interval of sensitivities
dat.enr <- data.frame(x = as.numeric(rownames(enr.obj)),
                      lower = enr.obj[,1],
                      upper = enr.obj[,3])

auc.lab <- paste0("AUC:", round(enr.auc$auc, 3), "\n (", round(enr.auc$ci[[1]], 3), "-", round(enr.auc$ci[[3]], 3), ")")

# turn into ggplot object
fig2d <- ggplot() +
  geom_rect(aes(xmin = 0, xmax = 1, ymin = 0, ymax = 1), fill = "grey86") +
  geom_abline(slope = 1, intercept = 1, linetype = "dashed", color = "grey30") +
  geom_ribbon(data = dat.enr, aes(x = x, ymin = lower, ymax = upper),
              fill = "#534582", alpha = 0.2) +
  geom_line(data = enr.df, aes(x = specificity, y = sensitivity), color = "#221C35", size = 1) +
  xlab("Specificity") +
  ylab("Sensitivity") +
  scale_x_reverse(lim=c(1, 0)) +
  annotate("text", x = 0.5, y = 0.5, fontface = "italic",
           vjust = 0.5, hjust = -0.5,
           size = 2,
           label = auc.lab) +
  theme(aspect.ratio = 1)
fig2d

# stat for AUC using DeLong, DeLong, and Clarke-Pearson (1988) method
z <- (enr.auc$auc - 0.5)/sqrt(var(enr.auc$ci))
2 * pt(-abs(z), df=Inf)
```


## Fig 2E. ENR score for both training and testing set
```{r}
enr.score.train <- data.frame(Group = mtdata$CMV, AnimalID = mtdata$AnimalID, cmv_score)
enr.score.test <- data.frame(Group = vacmtdata$Timepoint, AnimalID = vacmtdata$AnimalID, predicted)

enr.score <- rbind(enr.score.train, enr.score.test)
enr.score$Group <- factor(enr.score$Group, levels = c("NEG", "POS", "prev", "posv"))

fig2e <- ggplot(enr.score, aes(x = Group, y = s1, fill = Group)) +
  geom_boxplot(outlier.size = NA, alpha = 0.8) +
  geom_point(aes(color = Group), size = 2) +
  scale_fill_manual(values = set.color, limits = force, guide = "none") +
  scale_color_manual(values = set.color, limits = force, guide = "none") +
  scale_x_discrete(labels = c("NEG" = "RhCMV-",
                              "POS" = "RhCMV+",
                              "prev" = "Pre-vacc.",
                              "posv" = "Post-vacc.")) +
  labs(y = "CMV-microbial score") +
  theme(aspect.ratio = 1)
fig2e
```


## Fig 2F. PCA for significantly changed taxa in CMV38
```{r}
# get log transformed counts from CMV38
vac.sig <- subset_taxa(vac, Genus %in% sigbac)
vac.log <- transform_sample_counts(vac.sig, function(x) log2(x + 1))

logbac.vac <- data.frame(otu_table(vac.log))
colnames(logbac.vac) <- data.frame(as(tax_table(vac.log), "matrix"), stringsAsFactors = FALSE)$Genus

vac.pca <- prcomp(logbac.vac, scale. = TRUE)
vac.out <- cbind(vacmtdata, as.data.frame(vac.pca$x))
vac.percentage <- round(vac.pca$sdev^2 / sum(vac.pca$sdev^2) * 100, 2)

vac.eu <- dist(logbac.vac)

fig2f <- ggplot(vac.out, aes(x = PC1, y = PC2)) +
  geom_point(size = 2, alpha = 0.8, aes(color = Timepoint)) +
  labs(x = paste("PC1 (", vac.percentage[1], "%)", sep = ""),
       y = paste("PC2 (", vac.percentage[2], "%)", sep = ""),
       title = "Testing set",
       subtitle = "Pre- vs. post-vaccination") +
  scale_color_manual(name = "Vacc.", values = set.color, limits = force, labels = c("Pre-", "Post-")) +
  theme(legend.position = c(1, 1),
        legend.justification = c(1, 1),
        legend.background = element_rect(color = "black", size = 0.2),
        legend.spacing.y = unit(1, "pt"),
        legend.margin = margin(0, 2, 0, 2),
        aspect.ratio = 1,
        plot.title = element_text(size = 12, face = "bold", margin = margin(0, 0, 0, 0)),
        plot.subtitle = element_text(size = 8, face = "italic", margin = margin(0, 0, -6, 0), hjust = 1, vjust = 6)) +
  annotate("text",  x = -Inf, y = Inf, fontface = "italic",
           vjust = 1.4, hjust = -0.2, size = 2,
           label = paste0("P = ",
                    round(adonis(vac.eu ~ Timepoint, data = vac.out, method = "eu")$aov.tab$`Pr(>F)`[1],3)))
fig2f

# beta-disper p-value
round(permutest(betadisper(vac.eu, vacmtdata$Timepoint))$tab$'Pr(>F)'[1], 3)
```

## Put all plots together
```{r fig2, fig.show = "hide"}
fig2ab <- plot_grid(fig2a, fig2b, nrow = 1, labels = c("A", "B"), label_size = 16, hjust = -0.1, vjust = 1.1, rel_widths = c(2, 3), align = "h")
fig2ab <- plot_grid(fig2ab, get_legend(fig2.legend), ncol = 1, rel_heights = c(6, 1))

fig2abc <- plot_grid(fig2ab, fig2c, nrow = 1, labels = c("", "C"), label_size = 16, hjust = -0.1, vjust = 1.1, rel_widths = c(2, 1))

fig2def <- plot_grid(fig2d, fig2e, fig2f, labels = c("D", "E", "F"), label_size = 16, hjust = -0.1, vjust = 1.1, nrow = 1, align = "h")

plot_grid(fig2abc, fig2def, ncol = 1, align = "hv")

ggsave("211110_fig2.pdf", width = 9, height = 6.5, units = "in", device = "pdf")

sessionInfo()
```